Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-24 Manuscript/MPESubmission/Revision/response.tex

\begin{document}

%Add introduction (dear Dr blablalbal, here are our revisions)

%Resubmit the new version of the manuscript normaly but also add a pdf containing the track changes (highlithing in yellow the bit that have changed): "We are aware on the length of the manuscript. To facilitate the reviewers task, we attached to our revised manuscript, a pdf of the same manuscript but with the changes highlighted in yellow."

%TG: Add a section "Additional 'fancy' changes" (not fancy obviously)
% For clarity, we replaced the font of all the functions in R language in courier font (R, diversitree, etc...). \texttt{} - ACTUALLY DO THE CHANGE IF RELEVANT
%We are very grateful to both referees for their helpful and constructive comments, which we believe have helped us to significantly improve our paper. We have taken all of their comments on board, and respond to their points below.

@@

\end{enumerate}

\section{Reviewer 2:}

\begin{enumerate}
\item{\textcolor{blue}{\textbf{Compare the trees to the "true" tree:} I don't see how the "best" tree approach is helpful in this case. It adds an additional layer of complexity that can't be accounted for via simulation. The point of simulations is to compare empirical realities to the generating process, to see how much results diverge. I think the most helpful and straightforward thing to do is eliminate the "best" tree part of the MS, and compare the inferred trees directly to the real topology to quantify variance.}}

%TG: what does MS means? Manuscript?
%We do not agree blabalbalbal

@@

\item{\textcolor{blue}{\textbf{Better explanation of the missing data:} I found the explanation of the missing data strategy very difficult to follow. First, they oddly use the verb "to infer" for matrices; are they inferring matrices? Second, it's not clear anywhere that the authors actually removed the molecular data from the fossil taxa, though I assume they did. I hope the authors can re-write in simpler language ("explain like I'm 5") for easier comprehension.}}

%To do

\item{\textcolor{blue}{\textbf{$M_C$ is not missing data per se:} If Mc is just removing columns, then they are shrinking the total matrix. This is not missing data per se. They are instead comparing matrices of different sizes, and the quantity is "data present." Any matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. A matrix with 10,000 characters is tiny if 10,000,000 could be scored. This is semantic.}}

% TG: Ok i've changed it every where but personally I don't like it. It makes it even more messy and hard to read. However, because I didn't answered the first and last comment of this reviewer, I felt that at least 2/4 is a minimum.
We entirely agree with this comment and hope it will help to the global comprehension of the manuscript. We replaced the $M_C$ (missing characters) parameter by $N_C$ (number of characters with five states: 100, 90, 75, 50 and 25) in the revised manuscript, the appendices and the figures. We also slightly modified some explanations related to the change of missing characters to number of characters:

"$N_{C}$, the number of morphological characters for both living and fossil taxa: 100, 90, 75, 50 or 25 characters." line @@@

"When applying the $N_C$ parameter with the second parameter state (90 characters), however, the resulting matrix potentially had more than 10\% data missing.
Not that the $N_C$ parameter states can be seen in the same way as the $M_L$ and $M_F$ ones: the first state (100 characters) corresponds to 0\% less characters than the ``complete'' matrix, the second (90 characters), corresponds to 10\% less than the ``complete'' matrix, etc." line @@@

"To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $N_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\% and 100, 90, 75, 50, 25 characters), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 100; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $N_{C}$ = 100 etc. (see Figure 2 for more details)." line @@@

"Number of morphological characters ($N_{C}$)" line @@@ %Header in discussion

"Additionally, when both $M_{L}$ reach 50\% missing data and $N_{C}$ reach 50 characters, the matrix does not contain enough phylogenetic information for the fossil taxa to be placed with confidence in the tree (Figure 6)." line @@@

\item{\textcolor{blue}{\textbf{Specific variation of data per fossil:} More importantly, it doesn't seem like missing data varied among fossil taxa, unless I'm misreading. What we really want to know is what happens when some fossil taxa are fragmentary, as this is the most common empirical situation. Thus, there are morphological characters scored for some living taxa, some fossil taxa, and missing from others. It's unclear if this was simulated. It would be good if the authors could detail the case where, say, all living taxa had morphological data (0 Ml), the matrix was large (0 Mc), and Mf varied *among* fossil taxa, such that some fossils had 100\% coverage, and varied down to, say, 5\%. That's the crucial case here.}}

%Artifical fossilisation thing?
\end{enumerate}

### 2015-07-24 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
MC -> NC (in main + response)

\section{Materials and Methods}
To explore how missing data in the morphological partition of Total Evidence matrices influences tree topology, we used the following protocol (Figure 1):
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true'' tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete'' matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete'' matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). We call the resulting 125 matrices ``missing-data'' matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete'' matrix and from the 125 ``missing-data'' matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best'' tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data'' trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best'' tree to the ``missing-data'' trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

@@

\subsection{Removing data}
\label{Removing_data}
We modified the ``complete'' matrix to obtain matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \citep{sansomfossilization2013}. 
\item{$M_{C}$, the proportion of missing morphological characters for both living and fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters \citep[e.g][]{O'Leary08022013}.
\end{enumerate}

In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $M_C$ removes columns across both living and fossil taxa's data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.
We first applied the parameters $M_L$ and $M_F$ to the matrix and then applied the $M_C$ parameter.
Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix.
When applying the $M_C$ parameter with the same percentage (10\%), however, the resulting matrix potentially had more than 10\% data missing.

@@

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the ``complete'' matrix.
We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}.
We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete'' and 50 x 125 ``missing-data'' matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50).
This step was implemented to make sure that the differences we observed in topologies (see below) were due to the amount of missing data for each parameter ($M_L$, $M_F$ and $M_C$) and not simply to low branch support that is likely to lead to different topologies.
On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}.
Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes). We performed this procedure to increase the resolution of our resulting trees.

@@

\subsection{Comparing topologies}
We compared the topology of the ``missing-data'' trees to the ``best'' tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$ on tree topology.
We used the Robinson-Foulds distance \citep{RF1981} to assess the amount of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the amount of wildcard taxa \citep[i.e. taxa that frequently change position in different trees][]{kearneyfragmentary2002}. We used these two metrics because they illustrate two different aspects of tree topology (see Discussion) but also because their performance in measuring differences in topology is well described \citep{Kuhner04112014} and well implemented \citep{Bogdanowicz2012}.
We normalised both metrics using methods described in \citet{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa.
These metrics are described in detail below.

@@

\subsection{Testing the effects of the missing data parameters on topological recovery}
Finally, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on our ability to recover the ``best'' tree topology in a Total Evidence framework.
We also assessed the effect of our missing data parameters jointly with the effects of different tree inference and uncertainty methods (i.e. Maximum Likelihood, Bayesian consensus, Maximum Likelihood bootstrap trees and Bayesian posterior tree distribution).

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \citep{Bhattacharyya}.
The Bhattacharyya Coefficient is the probability of overlap between two distributions \citep{Bhattacharyya} (see Appendix B for calculation details).
Note that this is comparable to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing whole distributions not just their means.
To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Figure 2 for more details).
This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells).
We performed this procedure separately for each tree inference and uncertainty method.
When two combinations of missing data parameters have a similar ability to recover the ``best'' tree topology the Bhattacharyya Coefficient will be close to one.
Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.
Because of the difficulties to represent so many pairwise comparisons in a meaningful way, we summarized as a heat map of Bhattacharyya Coefficients (see Figure 6).
In this type of figures, parameters that have similar or different effects on recovering the ``best'' topology (rather good or bad effects) will be denoted by similar colour patches in the heat map representation of these comparisons (see Figure 6).

%\begin{figure}[h]
%\caption{{\bf Bhattacharyya Coefficient calculation outline 1.}
% A, B and C are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets metrics) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $M_{C}$ = 75\%). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two combinations of missing data parameters, for example, BC(A,B) is the probability of overlap between the distributions A and B. Note that this is similar to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
%\label{Fig_Bhattacharyya_Coefficients1}
%\end{figure}

@@

%\begin{figure}[h]
%\caption{{\bf Bhattacharyya Coefficient calculation outline 2.}
 %A and B are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets metrics) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $M_{C}$ = 75\%). \textbf{(x)} and \textbf{(y)} are two different tree inference methods (e.g. Maximum Likelihood or Bayesian). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two methods for the same combination of missing data parameters, for example, BC($A_{x}$,$A_{y}$) is the probability of overlap of the distribution A for methods $x$ and $y$. Note that this is similar to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
%\label{Fig_Bhattacharyya_Coefficients2} 
%\end{figure}

@@

\section{Results}
As the amount of missing data in the morphological part of the Total Evidence matrix increases, our ability to recover the ``best'' tree topology decreases, regardless of the missing data parameter ($M_{L}$, $M_{F}$ or $M_{C}$), the tree inference method (Maximum Likelihood or Bayesian) or the tree comparison metric used (Normalised Robinson-Foulds or Normalised Triplets metric).
Nonetheless, the different missing data parameters and tree inference methods do not affect the topology in the same way (Figure 4 and Figure 5).

%\begin{figure}[]
%\caption{{\bf The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (blue), Maximum Likelihood bootstrap trees (orange) and Bayesian posterior tree distributions (blue).}
 %The percentage of missing data for each parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds metric (upper row) and Normalised Triplets metric (lower row). The graph shows the modal value (points), and the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.}
%\label{Fig_Results-permeth_perparam} 
%\end{figure}

%\begin{figure}[]
%\caption{{\bf The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black) and Bayesian consensus trees (grey).}
 %The x axis shows the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $M_{C}$ (lower line). Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds metric (upper row) and Normalised Triplets metric (lower row). The graph shows the modal value (points), and the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.} 
%\label{Fig_Results-global_perparam}
%\end{figure}

\subsection{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the ``best'' tree topology decreases (Figure 4).
The Normalised Robinson-Foulds metric is always lower for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = 0.69, 0.48 and 0.66 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Figure 4; Tables C5, C6 and C7 in Appendix C). 
The Normalised Triplets metric, however, is similar between the Maximum Likelihood trees and the Bayesian consensus trees for all the parameters ($M_{L}$, $M_{F}$ and $M_{C}$) (median Bhattacharrya Coefficient = 0.84, 0.75 and 0.80 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Figure 4; Tables C5, C6 and C7 in Appendix C).

\subsection{Combined effect of missing data parameters}
As expected, our ability to recover the ``best'' tree topology is worst when each parameter contains the maximum amount of missing data (i.e. $M_{L}$ = 75\%, $M_{F}$ = 75\% and $M_{C}$ = 75\%), and best when there is no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; Figure 5; Tables C2, C3 and C4 in Appendix C).
Figure 6 shows the similarity of distributions of tree metrics in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (orange when the distributions overlap completely, Bhattacharyya Coefficient = 1, and blue when they do not, Bhattacharyya Coefficient = 0).

%\begin{figure}[]
%\caption{{\bf The effects of missing data on topological recovery using Bayesian consensus trees.}
% Both axes show the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $M_{C}$ (lower line). Topological recovery is represented by the probability of (A) Normalised Robinson-Foulds metric and (B) Normalised Triplets metric distributions overlapping with the ``best'' tree distribution, calculated using the Bhattacharyya Coefficient. The Bhattacharyya Coefficient values are indicated using a color gradient ranging from low probability of overlap in blue, to a high probability of overlap in orange.
%}
%\label{Fig_Results-paircomp_within}
%\end{figure}

Using both Normalised Robinson-Foulds and Normalised Triplets metrics from the Bayesian consensus trees, the parameter combination with no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%) is always the most dissimilar to all the other parameter combinations (thin deep blue line at the base of Figure 6).
The Normalised Robinson-Foulds metric (median Bhattacharrya coefficient = 0.79; blue regions in Figure 6A), however, displays more dissimilarities than the Normalised Triplets metric (median Bhattacharrya coefficient = 0.81; blue regions in Figure 6B).
The orange upper triangle in Figure 6A shows a high probability of overlap of the Normalised Robinson-Foulds metric for the trees with the $M_{L}$ parameter $\geq$ 50\% (Figure 6A).
There is no additional effect (i.e. no dissimilarities) of $M_{F}$ and $M_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
Additionally, when using the Normalised Robinson-Foulds metric, once $M_{L}$ $\geq$ 50\%, there is no additional effect of $M_{F}$ and $M_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
Likewise, once $M_{C}$ $\geq$ 50\%, there is no additional effect of $M_{L}$ and $M_{F}$ as denoted by the high probability of Normalised Robinson-Foulds metric overlap (horizontal orange stripes between the blue regions Figure 6A).

@@

\section{Conclusions}
Missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $M_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
Therefore we advise filling as many gaps in Total Evidence matrices as possible.
Because it can be difficult to do for fossil taxa, we recommend coding as many morphological characters as possible for living taxa.

>>>>

\section{Materials and Methods}
To explore how missing data in the morphological partition of Total Evidence matrices influences tree topology, we used the following protocol (Figure 1):
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the ``true'' tree) and used it to simulate a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the ``complete'' matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the ``complete'' matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the number of morphological characters ($N_{C}$). We call the resulting 125 matrices ``missing-data'' matrices.
\item{Estimating phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the ``complete'' matrix and from the 125 ``missing-data'' matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the ``best'' tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the ``missing-data'' trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the ``best'' tree to the ``missing-data'' trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $N_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

@@

\subsection{Removing data}
\label{Removing_data}
We modified the ``complete'' matrix to obtain matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \citep{sansomfossilization2013}. 
\item{$N_{C}$, the number of morphological characters for both living and fossil taxa: 100, 90, 75, 50 or 25 characters. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters \citep[e.g][]{O'Leary08022013}.
\end{enumerate}

In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $N_C$ removes columns across both living and fossil taxa's data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.
We first applied the parameters $M_L$ and $M_F$ to the matrix and then applied the $N_C$ parameter.
Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix.
When applying the $N_C$ parameter with the second parameter state (90 characters), however, the resulting matrix potentially had more than 10\% data missing.
Not that the $N_C$ parameter states can be seen in the same way as the $M_L$ and $M_F$ ones: the first state (100 characters) corresponds to 0\% less characters than the ``complete'' matrix, the second (90 characters), corresponds to 10\% less than the ``complete'' matrix, etc.

@@

To measure the support for each branch in our simulated phylogenies we first ran a fast bootstrap analysis (Lazy Sub-tree Rearrangement) with 500 replicates on the ``complete'' matrix.
We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}.
We repeated this selection until we obtained 50 sets of simulations (i.e. 50 ``complete'' and 50 x 125 ``missing-data'' matrices) with a relatively strong phylogenetic signal (median bootstrap $>$ 50).
This step was implemented to make sure that the differences we observed in topologies (see below) were due to the amount of missing data for each parameter ($M_L$, $M_F$ and $N_C$) and not simply to low branch support that is likely to lead to different topologies.
On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}.
Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes). We performed this procedure to increase the resolution of our resulting trees.

@@

\subsection{Comparing topologies}
We compared the topology of the ``missing-data'' trees to the ``best'' tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $N_{C}$ on tree topology.
We used the Robinson-Foulds distance \citep{RF1981} to assess the amount of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the amount of wildcard taxa \citep[i.e. taxa that frequently change position in different trees][]{kearneyfragmentary2002}. We used these two metrics because they illustrate two different aspects of tree topology (see Discussion) but also because their performance in measuring differences in topology is well described \citep{Kuhner04112014} and well implemented \citep{Bogdanowicz2012}.
We normalised both metrics using methods described in \citet{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa.
These metrics are described in detail below.

@@

\subsection{Testing the effects of the missing data parameters on topological recovery}
Finally, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$, $N_{C}$ and their interactions) on our ability to recover the ``best'' tree topology in a Total Evidence framework.
We also assessed the effect of our missing data parameters jointly with the effects of different tree inference and uncertainty methods (i.e. Maximum Likelihood, Bayesian consensus, Maximum Likelihood bootstrap trees and Bayesian posterior tree distribution).

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \citep{Bhattacharyya}.
The Bhattacharyya Coefficient is the probability of overlap between two distributions \citep{Bhattacharyya} (see Appendix B for calculation details).
Note that this is comparable to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing whole distributions not just their means.
To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $N_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\% and 100, 90, 75, 50, 25 characters), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 100; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $N_{C}$ = 100 etc. (see Figure 2 for more details).
This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells).
We performed this procedure separately for each tree inference and uncertainty method.
When two combinations of missing data parameters have a similar ability to recover the ``best'' tree topology the Bhattacharyya Coefficient will be close to one.
Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.
Because of the difficulties to represent so many pairwise comparisons in a meaningful way, we summarized as a heat map of Bhattacharyya Coefficients (see Figure 6).
In this type of figures, parameters that have similar or different effects on recovering the ``best'' topology (rather good or bad effects) will be denoted by similar colour patches in the heat map representation of these comparisons (see Figure 6).

%\begin{figure}[h]
%\caption{{\bf Bhattacharyya Coefficient calculation outline 1.}
% A, B and C are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets metrics) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $N_{C}$ = 75\%). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two combinations of missing data parameters, for example, BC(A,B) is the probability of overlap between the distributions A and B. Note that this is similar to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
%\label{Fig_Bhattacharyya_Coefficients1}
%\end{figure}

@@

%\begin{figure}[h]
%\caption{{\bf Bhattacharyya Coefficient calculation outline 2.}
 %A and B are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets metrics) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $N_{C}$ = 75\%). \textbf{(x)} and \textbf{(y)} are two different tree inference methods (e.g. Maximum Likelihood or Bayesian). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two methods for the same combination of missing data parameters, for example, BC($A_{x}$,$A_{y}$) is the probability of overlap of the distribution A for methods $x$ and $y$. Note that this is similar to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
%\label{Fig_Bhattacharyya_Coefficients2} 
%\end{figure}

@@

\section{Results}
As the amount of missing data in the morphological part of the Total Evidence matrix increases, our ability to recover the ``best'' tree topology decreases, regardless of the missing data parameter ($M_{L}$, $M_{F}$ or $N_{C}$), the tree inference method (Maximum Likelihood or Bayesian) or the tree comparison metric used (Normalised Robinson-Foulds or Normalised Triplets metric).
Nonetheless, the different missing data parameters and tree inference methods do not affect the topology in the same way (Figure 4 and Figure 5).

%\begin{figure}[]
%\caption{{\bf The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black), Bayesian consensus trees (blue), Maximum Likelihood bootstrap trees (orange) and Bayesian posterior tree distributions (blue).}
 %The percentage of missing data for each parameter ($M_{L}$, $M_{F}$ and $N_{C}$) is shown on the x axis. Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds metric (upper row) and Normalised Triplets metric (lower row). The graph shows the modal value (points), and the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.}
%\label{Fig_Results-permeth_perparam} 
%\end{figure}

%\begin{figure}[]
%\caption{{\bf The effects of increasing missing data on topological recovery using Maximum Likelihood trees (black) and Bayesian consensus trees (grey).}
 %The x axis shows the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $N_{C}$ (lower line). Topological recovery was measured using two different tree comparison metrics: Normalised Robinson-Foulds metric (upper row) and Normalised Triplets metric (lower row). The graph shows the modal value (points), and the 50\% (thick solid lines) and 95\% (thin dashed lines) confidence intervals of the distributions of the tree comparison metric for each missing data parameter and tree inference method.} 
%\label{Fig_Results-global_perparam}
%\end{figure}

\subsection{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the ``best'' tree topology decreases (Figure 4).
The Normalised Robinson-Foulds metric is always lower for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = 0.69, 0.48 and 0.66 for $M_{L}$, $M_{F}$ and $N_{C}$ respectively; Figure 4; Tables C5, C6 and C7 in Appendix C). 
The Normalised Triplets metric, however, is similar between the Maximum Likelihood trees and the Bayesian consensus trees for all the parameters ($M_{L}$, $M_{F}$ and $N_{C}$) (median Bhattacharrya Coefficient = 0.84, 0.75 and 0.80 for $M_{L}$, $M_{F}$ and $N_{C}$ respectively; Figure 4; Tables C5, C6 and C7 in Appendix C).

\subsection{Combined effect of missing data parameters}
As expected, our ability to recover the ``best'' tree topology is worst when each parameter contains the maximum amount of missing data (i.e. $M_{L}$ = 75\%, $M_{F}$ = 75\% and $N_{C}$ = 75\%), and best when there is no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 0\%; Figure 5; Tables C2, C3 and C4 in Appendix C).
Figure 6 shows the similarity of distributions of tree metrics in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (orange when the distributions overlap completely, Bhattacharyya Coefficient = 1, and blue when they do not, Bhattacharyya Coefficient = 0).

%\begin{figure}[]
%\caption{{\bf The effects of missing data on topological recovery using Bayesian consensus trees.}
% Both axes show the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $N_{C}$ (lower line). Topological recovery is represented by the probability of (A) Normalised Robinson-Foulds metric and (B) Normalised Triplets metric distributions overlapping with the ``best'' tree distribution, calculated using the Bhattacharyya Coefficient. The Bhattacharyya Coefficient values are indicated using a color gradient ranging from low probability of overlap in blue, to a high probability of overlap in orange.
%}
%\label{Fig_Results-paircomp_within}
%\end{figure}

Using both Normalised Robinson-Foulds and Normalised Triplets metrics from the Bayesian consensus trees, the parameter combination with no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 100) is always the most dissimilar to all the other parameter combinations (thin deep blue line at the base of Figure 6).
The Normalised Robinson-Foulds metric (median Bhattacharrya coefficient = 0.79; blue regions in Figure 6A), however, displays more dissimilarities than the Normalised Triplets metric (median Bhattacharrya coefficient = 0.81; blue regions in Figure 6B).
The orange upper triangle in Figure 6A shows a high probability of overlap of the Normalised Robinson-Foulds metric for the trees with the $M_{L}$ parameter $\geq$ 50\% (Figure 6A).
There is no additional effect (i.e. no dissimilarities) of $M_{F}$ and $N_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
Additionally, when using the Normalised Robinson-Foulds metric, once $M_{L}$ $\geq$ 50\%, there is no additional effect of $M_{F}$ and $N_{C}$, regardless of the amount of missing data in these parameters (Figure 6A).
Likewise, once $N_{C}$ $<$ 50, there is no additional effect of $M_{L}$ and $M_{F}$ as denoted by the high probability of Normalised Robinson-Foulds metric overlap (horizontal orange stripes between the blue regions Figure 6A).

@@

\section{Conclusions}
Missing data in Total Evidence matrices is not a problem for recovering the ``best'' tree topology as long as enough living and fossil taxa in the matrix have data for overlapping morphological characters.
When missing data increases in any of our missing data parameters ($M_{L}$, $M_{F}$ or $N_{C}$), it reduces support for the placement of fossil taxa and increases the displacement of wildcard taxa.
Therefore we advise filling as many gaps in Total Evidence matrices as possible.
Because it can be difficult to do for fossil taxa, we recommend coding as many morphological characters as possible for living taxa.
